Detection of Epistatic Interactions in High-Dimensional Genomic Data Using Random Forests

### 2025-05-02 manuscript.tex
added Supplementary_File_Rheumatoid_Data.Rmd, other changes to manuscript.tex

%% Use the option review to obtain double line spacing
\documentclass[authoryear,preprint,12pt,3p]{elsarticle}

@@

\usepackage{threeparttable}
\usepackage{url}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{makecell}
\usepackage{placeins}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage[figuresleft]{rotating}
%\usepackage{rotating}

@@

% \newcommand{\tikzmark}[1]{\tikz[overlay,remember picture] \node (#1) {};}
% \newcommand{\DrawBox}[4][]{%
%     \tikz[overlay,remember picture]{%
%         \coordinate (TopLeft)     at ($(#2)+(-0.2em,0.9em)$);
%         \coordinate (BottomRight) at ($(#3)+(0.2em,-0.3em)$);
%         %
%         \path (TopLeft); \pgfgetlastxy{\XCoord}{\IgnoreCoord};
%         \path (BottomRight); \pgfgetlastxy{\IgnoreCoord}{\YCoord};
%         \coordinate (LabelPoint) at ($(\XCoord,\YCoord)!0.5!(BottomRight)$);
%         %
%         \draw [red,#1] (TopLeft) rectangle (BottomRight);
%         \node [below, #1, fill=none, fill opacity=1] at (LabelPoint) {#4};
%     }
% }

@@

\tableofcontents 
\clearpage

@@

\author{Hawlader A. Al-Mamun\corref{cor1}\fnref{label2}}
\ead{hawlader.almamun@uwa.edu.au}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
\affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
  addressline={2 - 40 Clunies Ross Street}, 
  city={Acton},
  postcode={2601}, 
  state={ACT},
  country={Australia}}

@@

\author{Ross L. Tellam}
\affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
  addressline={306 Carmody Rd}, 
  city={St Lucia},
  postcode={4067}, 
  state={QLD},
  country={Australia}}

\author{Klara Verbyla}
\affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
  addressline={2 - 40 Clunies Ross Street}, 
  city={Acton},
  postcode={2601}, 
  state={ACT},
  country={Australia}}

@@

%\usepackage{caption}
%\usepackage{natbib}
%\usepackage[backend=biber,style=apa,defernumbers=true]{biblatex}
%\addbibresource{./manuscript.bib}
%\journal{Nuclear Physics B}

@@

Now, somewhat uniquely, we can move the contaminating distribution. All we need to do is fit a Random Forests with more
trees and we will observe more counts.
There is a trade-off here and as we are sampling a space with typically more than $10^{9}$ possible interactions,
we have to increase the number of trees a lot to get a meaningful difference in the mean of the contaminating
distribution. Because of this we resort to a threshholding stratey that is both computationally easier and controls the FDR rate by testing
each variable.

\subsection{threshholding the data}
In this case we have followed a simple strategy:
\begin{itemize}
\item threshold the observed count data. We do this a number of ways to do this (list them) and then take the median. In
  the case of the Rheumatoid data, this was 2 which gave us 222 interactions;
\item test each interaction with a linear model (see below).
\end{itemize}

@@

Notably, multiple genome wide association studies (GWAS) identified a subset of the interactive SNPs as associated with
various autoimmune diseases i.e., rs9273363, rheumatoid arthritis; type 1 diabetes; rs6457617, multiple sclerosis,
Graves disease, systemic sclerosis; rs615672, rheumatoid arthritis ("UCSC Genome Browser Home").
\marginpar{is UCSC Genome Browser Home a reference?}
The results fromusing the proposed method on the human rheumatoid arthritis data revealed a number of promising results supporting the
potential value of the proposed method for identifying epistatic interactions in case control data including the
ability to find positive and negative cis-acting interactions.

@@

\section{Acknowledgements}
The authors thank the anonymous reviewers for their valuable suggestions. This work is supported by CSIRO MLAI Future
Science Platform. This study makes use of data generated by the Wellcome Trust Case Control Consortium. A full list of
the investigators who contributed to the generation of the data is available from www.wtccc.org.uk. Funding for the
project was provided by the Wellcome Trust under award 076113.

@@

\bibliographystyle{elsarticle-harv}
\bibliography{./manuscript.bib}

>>>>

%% Use the option review to obtain double line spacing
\documentclass[preprint,12pt,3p]{elsarticle}
%%if you are using elsarticle.cls, do not use it with options
%%    \documentclass[authoryear]{elsarticle}
%% and the style
%% \bibliographystyle{elsarticle-num}
%% https://tex.stackexchange.com/questions/54480/package-natbib-error-bibliography-not-compatible-with-author-year-citations

@@

\usepackage{threeparttable}
\usepackage{url}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{makecell}
\usepackage{placeins}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage[figuresleft]{rotating}

@@

@@

@@

\author{Hawlader A. Al-Mamun\corref{cor1}\fnref{label2}}
\ead{halmamun@intergrain.com}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
\affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
  addressline={2 - 40 Clunies Ross Street}, 
  city={Acton},
  postcode={2601}, 
  state={ACT},
  country={Australia}}

@@

\author{Ross L. Tellam}
\affiliation{organization={CSIRO Agriculture and Food},%Department and Organization
  addressline={306 Carmody Rd}, 
  city={St Lucia},
  postcode={4067}, 
  state={QLD},
  country={Australia}}

\author{Klara Verbyla}
\affiliation{organization={The Center for Aquaculture Technologies},%Department and Organization
  addressline={8445 Camino Santa Fe, Suite 104}, 
  city={San Diego},
  postcode={92121}, 
  state={California},
  country={USA}}

@@

%\usepackage{capton}
%\usepackage{natbib}
%\usepackage[backend=biber,style=apa,defernumbers=true]{biblatex}
%\addbibresource{./manuscript.bib}
%\journal{Nuclear Physics B}

@@

Now, somewhat uniquely, we can move the contaminating distribution. All we need to do is fit a Random Forests with more
trees and we will observe more counts.  There is a trade-off here and as we are sampling a space with typically more
than $10^{9}$ possible interactions, we have to increase the number of trees a lot to get a meaningful difference in the
mean of the contaminating distribution. We have found the thresholding to give a larger set of significant variables
than the empirical Bayes method. However it controls the FDR rate by testing each variable and computationally easier.

\subsection{threshholding the data}
In this case we have followed a simple strategy:
\begin{itemize}
\item threshold the observed count data. We do this a number of ways to do this (breakpoints in the distribution,
  entropy tests and otehrs. See the supplementary informaton) and then take the median. In
  the case of the Rheumatoid data, this was 2 which gave us 222 interactions;
\item test each interaction with a linear model (see below).
\end{itemize}

@@

Notably, multiple genome wide association studies (GWAS) identified a subset of the interactive SNPs as associated with
various autoimmune diseases i.e., rs9273363, rheumatoid arthritis; type 1 diabetes; rs6457617, multiple sclerosis,
Graves disease, systemic sclerosis; rs615672, rheumatoid arthritis \cite[]{UCSC.2022}.
The results fromusing the proposed method on the human rheumatoid arthritis data revealed a number of promising results supporting the
potential value of the proposed method for identifying epistatic interactions in case control data including the
ability to find positive and negative cis-acting interactions.

@@

\section{Acknowledgements}
This work is supported by CSIRO MLAI Future
Science Platform. This study makes use of data generated by the Wellcome Trust Case Control Consortium. A full list of
the investigators who contributed to the generation of the data is available from www.wtccc.org.uk. Funding for the
project was provided by the Wellcome Trust under award 076113.

@@

\bibliographystyle{elsarticle-num}
\bibliography{./manuscript.bib}
